A large leading tail of atmospheric escape in the HAT-P-67 b system

### 2023-03-22 paper/paper1/ms.tex
Minor edits

\subsection{Size and geometry of outflow}
Helium absorption occurs predominantly before the planet is in-transit, indicating a \emph{leading} tail escaping the planet.  The excess occurs as early as 30\% of the orbital path.  That large fraction of an orbit translates to about 0.1 AU, 130 planetary radii, and well outside of the Roche Lobe, with its Hill radius of merely 2.6 planetary radii.  We are therefore primarily detecting gas that is entirely unbound from the planet.  The leading gas tail does not appear to accelerate, neither blue-shifting nor redshifting.   The bulk of the detectable unbound gas may therefore follow its own Keplerian trajectory at a slightly shorter period orbit, pulling away from the planet parallel to the orbital path.  Such a scenario may arise from preferential emission on the planet dayside, or other configurations that we discuss in Section \ref{secDiscuss}.

The out-of-transit excess exhibits approximately the same line width as the stellar lines, which are rotationally broadened by a $v\sin{i}$ of 30-35 km/s \citep{2017AJ....153..211Z}.  The large size of the gas stream overfills the transit chord, which means both the advancing blue side and receding red side of the rotating star are nearly continuously obscured in this spin-orbit-aligned system \citep{2017AJ....153..211Z}. The planet transits near the Stellar equator \citep{2017AJ....153..211Z}, meaning the transit chord covers about 10$\%$ of the projected stellar surface area.  Hypothetically a nearly 100$\%$ opaque optically thick stream of gas covering only the transit chord could reproduce the observed 10$\%$ absorption excess.  Alternatively---and more likely---an optically thin gas stream has some additional vertical extent perpendicular to the plane of orbit, which would act to overfill the entire projected stellar disk.  Assuming the gas has a spatially uniform optical depth of about 0.1 would reproduce the 10\% absorption excess.

@@

The predicted Helium ionization depends sensitively on the XUV flux \citep{2019ApJ...881..133O}.  Ideally we would have a measurement of the XUV spectrum of \hatp, but the limited facilities and distance of the source preclude these challenging observations.  Instead we constructed panchromatic X-ray to visible SEDs by scaling and stitching together synthetic spectra to provide a range of high and low $L_X/L_\mathrm{bol}$.  We chose a 6400 K PHOENIX \citep{husser13} solar metallicity photospheric spectral template with $\log{g}=3.5$, scaled to the solid angle seen by \hatpb.  The F7IV-V \object[tau Boo]{$\tau$ Boo} serves as the closest analog with synthetic x-ray coronal spectra available \citep{2011A&A...532A...6S}.  But \hatp resides closer to the Kraft break than does \object[tau Boo]{$\tau$ Boo}, and the move to hotter F stars may be associated with rapid changes in the stellar wind and other atmospheric changes leading to the prospect of much lower XUV luminosity for \hatp.  So there remains significant uncertainty about the applicability of the \object[tau Boo]{$\tau$ Boo} x-ray spectrum to \hatp.  We scaled the synthetic SED of \object[tau Boo]{$\tau$ Boo} retrieved from X-exoplanets \citep{2011A&A...532A...6S} so that the integral of ground-state ionizing photons ($\lambda<504\AA$) exhibited $L_X/L_\mathrm{bol} \in [10^{-7}, 10^{-5}]$.  Figure \ref{fig:XUV} shows two conceivable SEDs with a high and low radiation hardness.

@@

The SED shows two key features.   First the corona model does not extend all the way to 2600 $\AA$ photons capable of ionizing out of the 2 $^3$S Helium metastable state.  This apparent deficit probably does not matter because the F spectral type obtains most of its NUV photons ($\lambda\sim2600\AA$) from the Wein side of the photospheric spectrum, so the hardness of the radiation stems mostly from the assumptions of the corona flux level.  The wavelength-dependent cross section for absorption (not shown) is largest just blueward of the ionization thresholds, so the differences in the spectrum between 504 $\AA$ and 1000 $\AA$ have relatively little impact on the overall ionization out of the metastable state.

>>>>

\subsection{Size and geometry of outflow}
Helium absorption occurs predominantly before the planet is in-transit, indicating a \emph{leading} tail escaping the planet.  The excess occurs as early as 30\% of the orbital path.  That large fraction of an orbit translates to about 0.1 AU, 130 planetary radii, and well outside of the Roche Lobe, with its Hill radius of merely 2.6 planetary radii.  We are therefore primarily detecting gas that is entirely unbound from the planet.  The leading gas tail does not appear to accelerate, with only a modest blueshift.  The bulk of the detectable unbound gas may therefore follow its own Keplerian trajectory at a slightly shorter period orbit, pulling away from the planet parallel to the orbital path.  Such a scenario may arise from preferential emission on the planet dayside, or other configurations that we discuss in Section \ref{secDiscuss}.

The large size of the gas stream overfills the transit chord, which means both the advancing blue side and receding red side of the rotating star are nearly continuously obscured in this spin-orbit-aligned system \citep{2017AJ....153..211Z}. The planet transits near the Stellar equator \citep{2017AJ....153..211Z}, meaning the transit chord covers about 10$\%$ of the projected stellar surface area.  Hypothetically a nearly 100$\%$ opaque optically thick stream of gas covering only the transit chord could reproduce the observed 10$\%$ absorption excess.  Alternatively---and more likely---an optically thin gas stream has some additional vertical extent perpendicular to the plane of orbit, which would act to overfill the entire projected stellar disk.  Assuming the gas has a spatially uniform optical depth of about 0.1 would reproduce the 10\% absorption excess.

@@

The predicted Helium ionization depends sensitively on the XUV flux \citep{2019ApJ...881..133O}.  Ideally we would have a measurement of the XUV spectrum of \hatp, but the limited facilities and distance of the source preclude these challenging observations.  Instead we constructed panchromatic X-ray to visible SEDs by scaling and stitching together synthetic spectra to provide a range of high and low $L_X/L_\mathrm{bol}$.  We chose a 6400 K PHOENIX \citep{husser13} solar metallicity photospheric spectral template with $\log{g}=3.5$, scaled to the solid angle seen by \hatpb.  The F7IV-V \object[tau Boo]{$\tau$ Boo} serves as the closest analog with available synthetic x-ray coronal spectra \citep{2011A&A...532A...6S}.  But \hatp resides closer to the Kraft break than does \object[tau Boo]{$\tau$ Boo}, and the move to hotter F stars may be associated with weakening of the stellar wind and other atmospheric changes leading to the prospect of much lower XUV luminosity for HAT-P-67.  So there remains significant uncertainty about the applicability of the \object[tau Boo]{$\tau$ Boo} XUV spectrum to HAT-P-67.  We scaled the synthetic SED of \object[tau Boo]{$\tau$ Boo} retrieved from X-exoplanets \citep{2011A&A...532A...6S} so that the integral of ground-state ionizing photons ($\lambda<504\AA$) exhibited $L_X/L_\mathrm{bol} \in [10^{-7}, 10^{-5}]$.  Figure \ref{fig:XUV} shows two conceivable SEDs with a high and low radiation hardness.

@@

The SED shows that the corona model does not extend all the way to 2600 \AA~photons capable of ionizing out of the 2 $^3$S Helium metastable state.  This apparent deficit should be negligible, since the F spectral type obtains most of its NUV photons ($\lambda\sim2600$ \AA) from the Wein side of the photospheric spectrum, so the hardness of the radiation stems mostly from the assumptions of the corona flux level.  The wavelength-dependent cross section for absorption (not shown) is largest just blueward of the ionization thresholds, so the differences in the spectrum between 504 and 1000 \AA~ have relatively little impact on the overall ionization out of the metastable state.
